Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-27 icfp.tex
Fix in instrumented semantics
Type Preservation for Case K~xs

    \infer[]
    { K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T
    \\
    \Gamma, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
      K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]

@@

\begin{proof}
By structural induction on the instrumented operational semantics reductions.
\begin{description}
    \item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
\begin{tabbing}
    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\} : \tau, \Sigma$\\
    (2) $\G, \Theta\uparrow^\omega; \D, \Theta\uparrow^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\}, \Theta\downarrow^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta\downarrow^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\to e'\}} : \tau$\\
    Subcase $e$ is in WHNF (thus $= K~\ov{x_i}$)\\
    (5) $\G'; \D_a \Vdash K~\ov{x_i} : \s \gtrdot \ov{\D_i}$\` by subcase\\
    (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
    Subsubcase $\vdash_{alt}$ $Alt_{NWHNF}$\\
    (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
    (8) $\Theta' = \Theta$\` in this subcase because $(\Theta \mid K~\ov{x_i})\Downarrow(\Theta \mid K~\ov{x_i})$\\
    (9) $\G'; \ov{\D_i} \vdash K~\ov{x_i} : \s$\` by inv (5)\\
    (10) $\ov{\G'; \cdot \vdash x_i : \s_i}$\` by inv (9) for unrestricted $x$s\\
    (11) $\ov{\G'; \D_j \vdash x_j : \s_j}$\` by inv (9) for linear fields $x$s\\
    (12) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}] : \tau}$\` by $\D$-subst and $\omega$-subst lemmas (7,10,11)\\
    (13) $\G'; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\` by $\D$-subst lemma and $\ov{\D_i} = \D_a$ (9,12)\\
    (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by reconstructing\\
    (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
    Subsubcase $\vdash_{alt}$ $Alt_{0}$\\
    (7) $(\G', z{:}_{\D_a}\s)[\cdot/\D_a]_z, \ov{y_i{:}_\omega\sigma_i}; (\D_a, \D_b)[\cdot / \D_a] \vdash e' : \tau$\` by inv subcase\\
    (8) $\G', z{:}_{\cdot}, \ov{y_i{:}_\omega\sigma_i}; \D_b \vdash e' : \tau$\` by $z$ cannot occur in $\G'$\\
    (9) Since $\G';\D_a \Vdash K~\ov{x_i}$, in this subcase $\D_a = \cdot$\\
    (10) $\G', \z{:}_\cdot\sigma, \ov{y{:}_\omega\sigma}; \D_b \vdash e' : \tau$\` by (9)\\
    (11) $\G', \z{:}_\cdot\sigma; \D_b \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst (10)\\
    (12) $\G'; \D_b \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst\\
    (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
    (14) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
    Subcase $e$ is not in WHNF\\
    (5) $\G'; \D_a \vdash e : \s$\` by inv on case\\
    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
\end{tabbing}
\end{description}
\end{proof}

>>>>

    \infer[]
    { K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T
    \\
    \Gamma, z{:}_{\ov{y_j}}T, \ov{y_i :_\omega \sigma_i} ;\Delta , \ov{y_j :_1 \sigma_j} \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
      K~\ov{x_i}) : T , e' : \tau ,\Sigma \\
    \Gamma ;\Delta \vdash (\Theta' \mid e'\ov{[x_i/y_{i,j}]}[K~\ov{x_i}/z]
    ) \Downarrow (\Theta'' \mid v) :\tau , \Sigma }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}T~\{{\dots,K~\ov{y_{i,j}} \to e', {\dots} }\}})
    \Downarrow (\Theta'' \mid  v) : \tau , \Sigma}\\[1.5em]

@@

\begin{proof}
By structural induction on the instrumented operational semantics reductions.
\begin{description}
    \item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
\begin{tabbing}
    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\} : \tau, \Sigma$\\
    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\to e'\}} : \tau$\\
    Subcase $e$ is in WHNF (thus $= K~\ov{x_i}$)\\
    (5) $\G'; \D_a \Vdash K~\ov{x_i} : \s \gtrdot \ov{\D_i}$\` by subcase\\
    (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
    Subsubcase $\vdash_{alt}$ $Alt_{WHNF}$\\
    (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
    (8) $\Theta' = \Theta$\` in this subcase because $(\Theta \mid K~\ov{x_i})\Downarrow(\Theta \mid K~\ov{x_i})$\\
    (9) $\G'; \ov{\D_i} \vdash K~\ov{x_i} : \s$\` by inv (5)\\
    (10) $\ov{\G'; \cdot \vdash x_i : \s_i}$\` by inv (9) for unrestricted $x$s\\
    (11) $\ov{\G'; \D_j \vdash x_j : \s_j}$\` by inv (9) for linear fields $x$s\\
    (12) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}] : \tau}$\` by $\D$-subst and $\omega$-subst lemmas (7,10,11)\\
    (13) $\G'; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\` by $\D$-subst lemma and $\ov{\D_i} = \D_a$ (9,12)\\
    (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by reconstructing\\
    (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
    Subsubcase $\vdash_{alt}$ $Alt_{0}$\\
    (7) $(\G', z{:}_{\D_a}\s)[\cdot/\D_a]_z, \ov{y_i{:}_\omega\sigma_i}; (\D_a, \D_b)[\cdot / \D_a] \vdash e' : \tau$\` by inv subcase\\
    (8) $\G', z{:}_{\cdot}, \ov{y_i{:}_\omega\sigma_i}; \D_b \vdash e' : \tau$\` by $z$ cannot occur in $\G'$\\
    (9) Since $\G';\D_a \Vdash K~\ov{x_i}$, in this subcase $\D_a = \cdot$\\
    (10) $\G', \z{:}_\cdot\sigma, \ov{y{:}_\omega\sigma}; \D_b \vdash e' : \tau$\` by (9)\\
    (11) $\G', \z{:}_\cdot\sigma; \D_b \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst (10)\\
    (12) $\G'; \D_b \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst\\
    (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
    (14) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
    Subcase $e$ is not in WHNF\\
    (5) $\G'; \D_a \vdash e : \s$\` by inv on case\\
    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_{i,j}} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
    Subsubcase $\vdash_{alt}$ $Alt_{NWHNF}$\\
    (7) $\G', z{:}_{\irr{\D_a}}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \irr{\D_a}, \D_b \vdash e' : \tau$\` for \ov{\D_j = \lctag{\irr{\D_a}}{K_j}}, by inv.\\
    (8) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \ov{y_j{:}_1\s_j}, \D_b \vdash e' : \tau$\` by $\D$-bound to linear\\
    (9) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b, \ov{y_j{:}_1\s_j} \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,8)\\
    (10) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta \mid e : \s, e':\tau, \Sigma$\`by reconstructing\\
    (11) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta' \mid K~\ov{x_i} : \s, e':\tau, \Sigma$\`by induction hypothesis\\
    (12) $\G, \Theta'{\uparrow}^\omega, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j}, \Theta'{\uparrow}^1 \vdash (K~\ov{x_{i,j}}, \Theta'{\downarrow}^1_e, e', \Sigma) : \s \bigotimes(\Theta'\downarrow^1_\sigma) \otimes \tau \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (13) $Let~\G''=\G,\Theta'{\uparrow}^\omega~and~\D_c\D_d\D_4\D_5=\D,\Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
    (13a) $\G''; \D_c \vdash K~\ov{x_{i,j}} : \s$\\
    (13b) $\G'', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_d, \ov{y_j{:}_1\s_j} \vdash e' : \tau$\\
    (14) $\ov{\G''; \cdot \vdash x_i : \s_i}$\\
    (15) $\ov{\G''; \D_j \vdash x_j : \s_j}$\` for $\ov{D_j}=\D_c$, by inv\\
    (16) $\G'', z{:}_{\D_c}\s; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]} : \tau$\`by $1$-substs and $\omega$-substs lemmas\\
    (17) $\G''; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\`by $\D$-subst lemma (13a,16)\\
    (18) $\G; \D \vdash \Theta' \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\`by reconstructing\\
    (19) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
    Subsubcase $\vdash_{alt}$ $Alt_0$\\
    (7) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_b \vdash e' : \tau$\` by inv\\
    (8) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,7)\\
    (9) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
    (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
    (11) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
    (12) $Let~\G'' = \G,\Theta'{\uparrow}^\omega~and~\D_d,\D_4,\D_5 = \D, \Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
    (12a) $\G''; \cdot \vdash K~\ov{x_i} : \s$\`since $K$ has no linear fields in this subcase\\
    (12b) $\G'', z{:}_{\cdot}\s, \ov{y_i{:}_\omega\s_i}; \D_d \vdash e' : \tau$\\
    (13) $\ov{\G''; \cdot \vdash x_{i} : \s_i}$\`by inv (12a)\\
    (14) $\G'', z{:}_{\cdot}\s; \D_d \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst lemma\\
    (15) $\G''; \D_d \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst lemma\\
    (16) $\G; \D \vdash \Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by reconstructing\\
    (17) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
\end{tabbing}
\end{description}
\end{proof}
